Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.9)

### 2015-09-25 MSstats_manual_RMarkdown.Rmd
2015.09.24 : Meena - section 1,2, reference updates

---
title: "Statistical analysis of proteomics experiments with R and MSstats (v3.0.9)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "September 18, 2015"
output: pdf_document
---

# Prerequisites

@@

# DDA analysis with MSstats

@@

## 1. Preparing the data for MSstats input

@@

## 2. Processing the data

### 2.1 Normalizing and summarizing data with dataProcess

@@

#### 2.1.1 Default normalization and summarization options

@@

#### 2.1.2 Different summarization options

@@

### 2.2 Visualization of processed data

#### 2.2.1 Quality control and Normalization effects

@@

#### 2.2.2 Summarization effects

@@

## 3. Finding differentially abundant proteins across conditions

### 3.1 Comparing conditions with groupComparison

@@

### 3.2 Visualization of differentially abundant proteins

@@

# Convert skyline output to MSstats required format

@@

## 1. Preparing the data for MSstats input
Let's read the data.

@@

### 1.1 Preprocessing with DDA experiment from Skyline output

@@

### 2. Different options for Skyline in dataProcess
Additional options in `dataProcess` need to be specified for processing Skyline output are `skylineReport=TRUE` (remove `Truncated=TRUE` rows and handle intensity=0) and `censoredInt="0"` (use intensity=0 as censored to handle missing values and in this case, NA values are assumed as random missing).

@@

# Convert MaxQuant output to MSstats required input

@@

### 1. First, get protein ID information
```{r, eval=T}
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

### 2. Read in annotation including condition and biological replicates: annotation.csv
```{r, eval=T}
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

### 3. Read in MaxQuant file: evidence.txt
```{r, eval=F}
infile <- read.table("evidence.txt", sep = "\t", header = TRUE)
```

@@

### 4. Reformat for MSstats required input
```{r, eval=T, warning=F}
# check options for converting format
?MaxQtoMSstatsFormat
```

@@

# now 'msstats.raw' is ready for MSstats
head(msstats.raw)
```

>>>>

---
title: "Protein significance analysis of mass spectrometry-based proteomics experiments with R and MSstats (v3.0.9)"
author: "Meena Choi, Tsung-Heng Tsai & Erik Verschueren"
date: "September 24, 2015"
output: pdf_document
bibliography: bibliography.bib
---

# 1. Statistical relative protein quantification: SRM, DDA and DIA experiments

MSstats is an open-source R-based package for statistical relative quantification of peptides and proteins in mass spectrometry-based proteomic experiments. This document describes `MSstats`, the most recent version of the package, and its use through the command line.

## Applicability
`MSstats` version 3.0 and above is applicable to multiple types of sample preparation, including label-free workflows, workflows that use stable isotope labeled reference proteins and peptides, and workflows that use fractionation. It is applicable to targeted Selected Reaction Monitoring (SRM), Data-Dependent Acquisition (DDA or shotgun), and Data-Independent Acquisition (DIA or SWATH-MS). It is applicable to experiments that make arbitrary complex comparisons of experimental conditions or times.

`MSstats` is currently not applicable to experiments that compare multiple metabolically labeled endogenous samples within a same run. It is not applicable to experiments with iTRAQ labeling. These experiments will be supported in the future.

## Statistical functionalities

`MSstats` version 3.0 and above performs three analysis steps. The first step, *data processing and visualization*, transforms and normalizes the intensities of the peaks, and generates workflow-specific and customizable numeric summaries for data visualization and quality control.

The second step, *statistical modeling and inference*, automatically detects the experimental design (e.g. group comparison, paired design or time course, presence of labeled reference peptides or proteins) from the data. It then reflects the experimental design, the type of spectral acquisition strategy, and the scope of conclusions (e.g. restricted to the subjects, or expanded to the underlying populations), and fits an appropriate linear mixed model by means of `lm` and `lmer` functionalities in R. The model is used to detect differentially abundant proteins or peptides, or to summarize the protein or peptide abundance in a single biological replicate or condition (that can be used, e.g. as input to clustering or classification).

The third step, *statistical experimental design*, views the dataset being analyzed as a pilot study of a future experiment, utilizes the variance components of the current datasets, and calculates the minimal number of replicates necessary in the future experiment to achieve a pre-specified statistical power.

## Interoperability with existing computational tools

`MSstats` takes as input data in a tabular .csv format, which can be generated by any spectral processing tool such as SuperHirn[@Mueller:2007fo], MaxQuant[@Cox:2008ir], Progenesis, MultiQuant, OpenMS[@Sturm:2008eu] or OpenSWATH.

For statistics experts, `MSstats` 3.0 and above satisfies the interoperability requirements of Bioconductor, and takes as input data in the `MSnSet` format[@Gatto:2012tj]. The command line-based workflow is partitioned into a series of independent steps, that facilitate the development and testing of alternative statistical approaches. It complies with the maintenance and documentation requirements of Bioconductor.

Finally, `MSstats` 3.0 and above is available as an external tool within Skyline [@MacLean:2010hd]. The external tool support within Skyline manages `MSstats` installation, point-and-click execution, parameter collection in Windows forms and output display.  Skyline manages the annotations of the experimental design, and the processing of raw data. It outputs a custom report, that is fed as a single stream input into `MSstats`.  This design buffers proteomics users from the details of the R implementation, while enabling rigorous statistical modeling.

## Availability

`MSstats` is available under the Artistic-2.0 license at [msstats.org](msstats.org). `MSstats` as an external tool is available at [http://proteome.gs.washington.edu/software/Skyline/tools.html](http://proteome.gs.washington.edu/software/Skyline/tools.html). `MSstats` is now also available in [Bioconductor](http:\\www.bioconductor.org). The most recent version of the package is available at [msstats.org](msstats.org) or [MSstats github](https://github.com/MeenaChoi/MSstats). We suggest to use that if possible. The versioning of the main package is updated several times a year, to synchronise with the Bioconductor release.

## Overview of the functionalities

![](/Users/Meena/Dropbox/MSstats_GitHub_document/MSstatsOverview.pdf)

## Troubleshooting

To help troubleshoot potential problems with installation or functionalities of `MSstats`, a progress report is generated in a separate log file *msstats.log*. The file includes information on the R session (R version, loaded software libraries), options selected by the user, checks of successful completion of intermediate analysis steps, and warning messages. If the analysis produces an error, the file contains suggestions for possible reasons for the errors. If a file with this name already exists in working directory, a suffix with a number will be appended to the file name. In this way a record of all the analyses is kept. Please see the file `KnownIssues-Skyline-MSstatsV2.1.6.pdf` on the `Installation` page of `msstats.org` for a list of known issues and possible solutions for installation problem of MSstats external tool in Skyline

# 2. Allowable data formats

## 2.1 SRM with stable isotope labeled reference peptides

### 2.1.1 10-column format

`MSstats` performs statistical analysis steps, that follow peak identification and quantitation. Therefore, input to MSstats is the output of other software tools (such as Skyline or MultiQuant) that read raw spectral files and identify and quantify spectral peaks. The preferred structure of data for use in `MSstats` is  a .csv file in a `long` format with 10 columns. The file should contain the following variables: `ProteinName`, `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, `ProductCharge`, `IsotopeLabelType`, `Condition`, ` BioReplicate`, `Run`, `Intensity`.  The variable names are fixed, but are not case-sensitive. This required input data is generated automatically when using `MSstats` report format in Skyline.

(a) `ProteinName`: This column needs information about Protein id. Statistical analysis will be done separately for each unique label in this column. For peptide-level modeling and analysis, use peptide id in this column.

(b)-(e) `PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge`: The combination of these 4 columns defines a *feature* of a protein (in SRM experiments, it is a transition that is identified and quantified across runs). If the information for one or several of these columns is not available, please do not discard these columns but use a single fixed value across the entire dataset. For example, if the original raw data does not contain the information of `ProductCharge`, assign the value 0 to the entries in the column `ProductCharge` for the entire dataset. If the peptide sequences should be distinguished based on post-translational modifications, this column can be renamed to `PeptideModifiedSequence`. For example, this allows us to use the `PeptideModifiedSequence` column from the Skyline report.

(f) `IsotopeLabelType`: This column indicates whether this measurement is based on the endogenous peptides (use `L`) or labeled reference peptides (use `H`).

(g) `Condition`: For group comparison experiments, this column indicates groups of interest (such as `Disease` or `Control`). For time-course experiments, this column indicates time points (such as `T1`, `T2`, etc). If the experimental design contains both distinct groups of subjects and multiple time points per subject, this column should indicate a combination of these values (such as `Disease_T1`, `Disease_T2`, `Control_T1`, `Control_T2`, etc).

(h) `BioReplicate`: This column should contain a unique identifier for each biological replicate in the experiment. For example, in a clinical proteomic investigation this should be a unique patient id. Patients from distinct groups should have distinct ids. `MSstats` does not require the presence of technical replicates in the experiment. If the technical replicates are present, all samples or runs from a same biological replicate should have a same id. `MSstats` automatically detects the presence of technical replicates and accounts for them in the model-based analysis.

(i) `Run`: This column contains the identifier of a mass spectrometry run. Each mass spectrometry run should have a unique identifier, regardless of the origin of the biological sample. In SRM experiments, if all the transitions of a  biological or a technical replicate are split into multiple `methods` due to the technical limitations, each method should have a separate identifier. When processed by Skyline, distinct values of runs correspond to distinct input file names. It is possible to use the actual input file names as values in the column `Run`.

(j) `Intensity`: This column should contain the quantified signal of a feature in a run without any transformation (in particular, no logarithm transform). The signals can be quantified as the peak height or the peak of area under curve. Any other quantitative representation of abundance can also be used.

An example of an input dataset is shown. Example dataset rom an SRM experiment with stable isotope labeled reference peptides. The dataset is stored in a .csv file in a `long` format.  Each row corresponds to a single intensity. More details on assigning the values of `Condition, BioReplicate` and `Run`, depending on the structure of the experimental design, are given below.

![](/Users/Meena/Dropbox/MSstats_GitHub_document/requiredInput.jpg)

### 2.1.2 Assigning the values of `Condition, BioReplicate` and `Run`

The values of `Condition, BioReplicate, Run` depend on the design of the specific experiment.

#### 1) Group comparison

In a group comparison design, the conditions (e.g., disease states) are profiled across non-overlapping sets of biological replicates (i.e. subjects). In this example there are 2 conditions, Disease and Control (in general the number of conditions can vary). There are 3 subjects (i.e. biological replicates) per condition (in general an equal number of replicates per condition is not required). Each subject has 2 technical replicate runs (in general technical replicates are not required, and their number per sample may vary). Overall, in this example there are $2 \times 3 \times 2 = 12$ mass spectrometry runs.

Table below shows the values of the columns `Condition, BioReplicate` and `Run` for this situation. It is important to note two things. First, the order of subjects and conditions in the experiment should be randomized, and run id does not need to represent the order of spectral acquisition. Second, the values of the columns are repeated for every quantified transition. For example, if in each run the experiment quantifies 50 endogenous transitions and 50 labeled reference counterparts, then the input file has $12 \times 50 \times 2 = 1200$ lines. When a feature intensity is missing in a run, the data structure should contain a separate row for each missing value. The rows should include all the information (from `ProteinName` to `Run`), and indicate missing intensities with `NA`.

`Condition` | `BioReplicate` |  `Run` 
------------|----------------|---------
Disease	|		Subject1	|		1
Disease	|		Subject1	|		2
Disease	|		Subject2	|		3
Disease	|		Subject2	|		4
Disease	|		Subject3	|		5
Disease	|		Subject3	|		6
Control |		Subject4	|		7
Control	|		Subject4	|		8
Control	|		Subject5	|		9
Control	|		Subject5	|		10
Control	|		Subject6	|		11
Control	|		Subject6	|		12

#### 2) Time course

The important feature of a time course experimental design is that a same subject (i. e. biological replicate) is repetitively measured across multiple time points. In this example there are 2 time points, Time1 and Time2 (in general the number of times can vary). There are 4 subjects (i.e. biological replicates) measured across times (in general an equal number of times per replicate is not required). There are no technical replicates (in general the number of technical replicates per sample may vary). Overall, in this example there are $2  \times 4 \times 1 = 8$ mass spectrometry runs.

Table below shows the values of the columns `Condition, BioReplicate` and `Run` for this situation. Comments on the order of the runs, on the number of lines in the input data structure, and on the handling of missing peak intensities are as in the group comparison design.

`Condition` |   `BioReplicate`|	 	`Run`
------------|----------------|---------
Time1	|		Subject1	|		1
Time2	|		Subject1	|		2
Time1	|		Subject2	|		3
Time2	|		Subject2	|		4
Time1	|		Subject3	|		5
Time2	|		Subject3	|		6
Time1	|		Subject4	|		7
Time2	|		Subject4	|		8

#### 3) Paired design

Another frequently used experimental design is a *pared design*, where measurements from multiple conditions (such as healthy biopsy and disease biopsy) are taken from a same subject. The statistical model for this experimental design is the same as in the time course experiment, however the values in the columns of the input data may have a different appearence. In this example there are 2 subjects, PatientA and PatientB (in general the number of patients can vary). There are two conditions per subject, BiopsyHealthy and BiopsyTumor (in general the number of conditions per subject can exceed two). In this example there are 3 technical replicates of each type (in this example, the technical replicates are biopsies; in general these can also be replicate sample preparations or replicate mass spectrometry runs). Overall, in this example there are $2  \times 2 \times 3 = 12$ mass spectrometry runs.

Table below shows the values of the columns `Condition, BioReplicate` and `Run` for this situation. Comments on the order of the runs, on the number of lines in the input data structure, and on the handling of missing peak intensities are as in the group comparison design.

`Condition` |  `BioReplicate` |   	`Run` 
------------|----------------|---------
BiopsyHealthy |		PatientA	|		1
BiopsyHealthy	|		PatientA	|		2
BiopsyHealthy	|	PatientA	|		3
BiopsyTumor	|		PatientA	|		4
BiopsyTumor	|		PatientA	|		5
BiopsyTumor	|		PatientA	|		6
BiopsyHealthy	|		PatientB	|		7
BiopsyHealthy |		PatientB	|		8
BiopsyHealthy	|		PatientB  |		9
BiopsyTumor	|		PatientB	|		10
BiopsyTumor	|		PatientB	|		11
BiopsyTumor	|		PatientB	|		12

### 2.1.3 `MSnSet` format

`MSstats` also allows data to be in the format of `MSnSet`, which is consistent with the requirements of Bioconductor. The `MSnSet` format has several components, of which the most commonly accessed are ` assayData`, `phenoData`, and `featureData`.  `assayData` is a matrix of intensities, where each row corresponds a transition, and the columns correspond to sample ids. `phenoData` contains columns that describe the biological samples, conditions in the experiment.  `featureData` contains columns describing the peptide features, such as the name or id of the underlying protein and information of features.

If the data are stored in the format `expressionSet`, group labels information is required. If more than one variable is listed in the argument `group`, then a concatenated variable is created based on all of the specified `group` variables. The remaining information (peptide feature ids, biological replicate ids, and abundance) can be extracted from the rows and columns of `featureData` and `phenoData`, or the users can assign them based on their experimental design.

## 2.2 Label-free DDA

For label-free DDA experiments the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. In DDA experiments spectral features are defined as peptide ions, which are identified and quantified across runs. Since for label-free DDA experiments some of the columns `PeptideSequence`, `PrecursorCharge`, `FragmentIon`, and `ProductCharge` are not relevant, these columns will have a constant fixed value (such as `NA`) across the entire dataset. Furthermore, the column `IsotopeLabelType` will be set to `L` for the entire dataset.

![](/Users/Meena/Dropbox/MSstats_GitHub_document/required_LCMS.png)

## 2.2 Label-free DIA

For label-free DIA experiments, the required input is the 10-column format, the same as described in section 2.1 for SRM experiments. The values of the required columns can be extracted from the output of signal processing software such as Skyline or OpenSWATH.  By default, the combination of the values in the columns `PeptideSequence, PrecursorCharge, FragmentIon, ProductCharge` uniquely identifies each spectral feature (i.e. a fragment ion identified and quantified across multiple runs). If the signal processing software does not provide the information on some of these columns but provides a unique feature identifier, it is possible to use this unique identifier instead of one of these columns. Furthermore, the column `IsotopeLabelType` is set to `L` for the entire dataset.

An example dataset is shown below. In this example, feature id generated by OpenSWATH is used instead of `ProductCharge` to uniquely characterize each feature.

![](/Users/Meena/Dropbox/MSstats_GitHub_document/required_DIA.png)

# 3. Prerequisites

@@

# 4. DDA analysis with MSstats

@@

## 4.1 Preparing the data for MSstats input

@@

## 4.2 Processing the data

### 4.2.1 Normalizing and summarizing data with dataProcess

@@

#### 4.2.1.1 Default normalization and summarization options

@@

#### 4.2.1.2 Different summarization options

@@

### 4.2.2 Visualization of processed data

#### 4.2.2.1 Quality control and Normalization effects

@@

#### 4.2.2.2 Summarization effects

@@

## 4.3 Finding differentially abundant proteins across conditions

### 4.3.1 Comparing conditions with groupComparison

@@

### 4.3.2 Visualization of differentially abundant proteins

@@

# 5. Convert skyline output to MSstats required format

@@

## 5.1 Preparing the data for MSstats input
Let's read the data.

@@

### 5.2 Preprocessing with DDA experiment from Skyline output

@@

### 5.3 Different options for Skyline in dataProcess
Additional options in `dataProcess` need to be specified for processing Skyline output are `skylineReport=TRUE` (remove `Truncated=TRUE` rows and handle intensity=0) and `censoredInt="0"` (use intensity=0 as censored to handle missing values and in this case, NA values are assumed as random missing).

@@

# 6. Convert MaxQuant output to MSstats required input

@@

```{r, eval=T}
## 1. First, get protein ID information
proteinGroups <- read.table("DDA2014_proteinGroups.txt", sep = "\t", header = TRUE)
```

```{r, eval=T}
## 2. Read in annotation including condition and biological replicates: annotation.csv
annot <- read.csv("DDA2014_annotation.csv", header = TRUE)
```

```{r, eval=F}
## 3. Read in MaxQuant file: evidence.txt
infile <- read.table("evidence.txt", sep = "\t", header = TRUE)
```

@@

```{r, eval=T, warning=F}
## 4. Reformat for MSstats required input
##    check options for converting format
?MaxQtoMSstatsFormat
```

@@

## now 'msstats.raw' is ready for MSstats
head(msstats.raw)
```

***
# Reference
